---
title: "2024 - Day 1: Historian Hysteria"
format:
  html:
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#AB63BD"
---

> You haven't even left yet and the group of Elvish Senior Historians has already hit a problem: their list of locations to check is currently empty. Eventually, someone decides that the best place to check first would be the Chief Historian's office.
> 
> Upon pouring into the office, everyone confirms that the Chief Historian is indeed nowhere to be found. Instead, the Elves discover an assortment of notes and lists of historically significant locations! This seems to be the planning the Chief Historian was doing before he left. Perhaps these notes can be used to determine which locations to search?
> 
> Throughout the Chief's office, the historically significant locations are listed not by name but by a unique number called the location ID. To make sure they don't miss anything, The Historians split into two groups, each searching the office and trying to create their own complete list of location IDs.
> 
> There's just one problem: by holding the two lists up side by side (your puzzle input), it quickly becomes clear that the lists aren't very similar. Maybe you can help The Historians reconcile their lists?

## Set-Up

```{r setup}
input <- read.table("input.txt")
```

## Part 1

> Maybe the lists are only off by a small amount! To find out, pair up the numbers and measure how far apart they are. Pair up the smallest number in the left list with the smallest number in the right list, then the second-smallest left number with the second-smallest right number, and so on.
>
> Your actual left and right lists contain many location IDs. What is the total distance between your lists?

```{r part_1}
(sort(input$V1) - sort(input$V2)) |> abs() |> sum()
```

## Part 2

> Your analysis only confirmed what everyone feared: the two lists of location IDs are indeed very different.
> 
> Or are they?
> 
> The Historians can't agree on which group made the mistakes or how to read most of the Chief's handwriting, but in the commotion you notice an interesting detail: a lot of location IDs appear in both lists! Maybe the other numbers aren't location IDs at all but rather misinterpreted handwriting.
> 
> This time, you'll need to figure out exactly how often each number from the left list appears in the right list. Calculate a total similarity score by adding up each number in the left list after multiplying it by the number of times that number appears in the right list.
>
> Once again consider your left and right lists. What is their similarity score?

```{r part_2}
(sapply(input$V1, \(x) sum(x == input$V2)) * input$V1) |> sum()
```
