---
title: "2022 - Day 6"
format:
  html:
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#AB63BD"
---

> The expedition can depart as soon as the final supplies have been unloaded from the ships. Supplies are stored in stacks of marked crates, but because the needed supplies are buried under many other crates, the crates need to be rearranged.
>
> The ship has a giant cargo crane capable of moving crates between stacks. To ensure none of the crates get crushed or fall over, the crane operator will rearrange them in a series of carefully-planned steps. After the crates are rearranged, the desired crates will be at the top of each stack.
>
> The Elves don't want to interrupt the crane operator during this delicate procedure, but they forgot to ask her which crate will end up where, and they want to be ready to unload them as soon as possible so they can embark.

## Set-Up

Today's has been the most complicated format, with two bits of code to read in, the first of which isn't structured nicely to read in with something like `read.table`.

```
        [Q] [B]         [H]
    [F] [W] [D] [Q]     [S]
    [D] [C] [N] [S] [G] [F]
    [R] [D] [L] [C] [N] [Q]     [R]
[V] [W] [L] [M] [P] [S] [M]     [M]
[J] [B] [F] [P] [B] [B] [P] [F] [F]
[B] [V] [G] [J] [N] [D] [B] [L] [V]
[D] [P] [R] [W] [H] [R] [Z] [W] [S]
 1   2   3   4   5   6   7   8   9
```

With some manipulation. `read.fwf` was the best solution I could find, taking each four characters and extracting the letter from each set of characters.
 
```{r setup}
blocks <- read.fwf("input.txt", widths = rep(4, 9), n = 8) |>
  lapply(\(x) rev(gsub("\\W", "", x))) |>
  lapply(\(x) x[!x %in% c("", NA)])
str(blocks)

operations <- read.table(
  "input.txt", sep = " ", skip = 10,
  col.names = c("x1", "move", "x2", "from", "x3", "to")
)
head(operations)
```

## Part 1

> After the rearrangement procedure completes, what crate ends up on top of each stack?

```{r part_1}
blocks_p1 <- blocks
for (i in seq(nrow(operations))) {
  blocks_p1[[operations$to[i]]] <- c(
    blocks_p1[[operations$to[i]]],
    rev(tail(blocks_p1[[operations$from[i]]], operations$move[i]))
  )
  blocks_p1[[operations$from[i]]] <- head(blocks_p1[[operations$from[i]]], -operations$move[i])
}
vapply(blocks_p1, tail, character(1L), n = 1) |> paste(collapse = "")
```

## Part 2

> As you watch the crane operator expertly rearrange the crates, you notice the process isn't following your prediction.
>
> Some mud was covering the writing on the side of the crane, and you quickly wipe it away. The crane isn't a CrateMover 9000 - it's a CrateMover 9001.
>
> The CrateMover 9001 is notable for many new and exciting features: air conditioning, leather seats, an extra cup holder, and the ability to pick up and move multiple crates at once.

Sometimes it is as simple as removing a `rev` from the previous solution...

```{r part_2}
for (i in seq(nrow(operations))) {
  blocks[[operations$to[i]]] <- c(
    blocks[[operations$to[i]]],
    tail(blocks[[operations$from[i]]], operations$move[i])
  )
  blocks[[operations$from[i]]] <- head(blocks[[operations$from[i]]], -operations$move[i])
}
vapply(blocks, tail, character(1L), n = 1) |> paste(collapse = "")
```
