---
title: "2023 - Day 3: Gear Ratios"
format:
  html:
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#AB63BD"
---

> You and the Elf eventually reach a gondola lift station; he says the gondola lift will take you up to the water source, but this is as far as he can bring you. You go inside.
>
> It doesn't take long to find the gondolas, but there seems to be a problem: they're not moving.
>
> "Aaah!"
>
> You turn around to see a slightly-greasy Elf with a wrench and a look of surprise. "Sorry, I wasn't expecting anyone! The gondola lift isn't working right now; it'll still be a while before I can fix it." You offer to help.
>
> The engineer explains that an engine part seems to be missing from the engine, but nobody can figure out which one. If you can add up all the part numbers in the engine schematic, it should be easy to work out which part is missing.

## Set-Up

```{r setup}
schema_lines <- readLines("input.txt")
schema <- do.call(rbind, strsplit(schema_lines, ""))
dims <- dim(schema)
```

## Part 1

> The engine schematic (your puzzle input) consists of a visual representation of the engine. There are lots of numbers and symbols you don't really understand, but apparently any number adjacent to a symbol, even diagonally, is a "part number" and should be included in your sum. (Periods (.) do not count as a symbol.)
>
> What is the sum of all of the part numbers in the engine schematic?

Couldn't think of a nice way to get two numbers when in the one before and one after the column position, so just went back to regular expressions to extract all numbers in a row and see if any of them were near the positions of special characters. `arr.ind` makes finding the position of logical matrices much easier.

TIL `==` maintains a matrix, `%in%` doesn't, hence the for loops checking special characters (and numbers in part 2).

```{r part_1}
special_chars <- unique(grep("[^0-9\\.]", schema, value = TRUE))
special_locs <- schema == special_chars[1]
for (i in special_chars) special_locs <- special_locs | schema == i

special_area <- which(special_locs, arr.ind = TRUE) |>
  apply(1, simplify = FALSE, \(x) {
    dat <- cbind(
      row = rep((x[1] - 1):(x[1] + 1), each = 3),
      col = rep((x[2] - 1):(x[2] + 1), times = 3)
    )
    # Removing cases outside the matrix
    dat[dat[, 1] >= 1 & dat[, 2] >= 1 & dat[, 1] <= dims[1] & dat[, 2] <= dims[2], ]
  }) |>
  do.call(what = rbind)

inner_nums <- numeric()
for (i in seq(dims[1])) {
  special_cols <- special_area[special_area[, 1] == i, "col"]
  number_locs <- gregexpr("[0-9]+", schema_lines[i])[[1]]
  if (number_locs[1] == -1) next
  for (j in seq_along(number_locs)) {
    start <- number_locs[j]
    area <- start + seq(attr(number_locs, "match.length")[j]) - 1
    if (any(area %in% special_cols)) {
      inner_nums <- c(inner_nums, as.numeric(substr(schema_lines[i], area, max(area))))
    }
  }
}

sum(inner_nums)
```

## Part 2

> The engineer finds the missing part and installs it in the engine! As the engine springs to life, you jump in the closest gondola, finally ready to ascend to the water source.
>
> You don't seem to be going very fast, though. Maybe something is still wrong? Fortunately, the gondola has a phone labeled "help", so you pick it up and the engineer answers.
>
> Before you can explain the situation, she suggests that you look out the window. There stands the engineer, holding a phone in one hand and waving with the other. You're going so slowly that you haven't even left the station. You exit the gondola.
>
> The missing part wasn't the only issue - one of the gears in the engine is wrong. A gear is any * symbol that is adjacent to exactly two part numbers. Its gear ratio is the result of multiplying those two numbers together.
>
> This time, you need to find the gear ratio of every gear and add them all up so that the engineer can figure out which gear needs to be replaced.
>
> What is the sum of all of the gear ratios in your engine schematic?

Thankfully no edge asterisks in this input, so didn't have to worry too much edge cases, but am sure there is a much nicer way to check for numbers than the way I've been using `sub` and (more) regular expressions.

```{r part_2}
special_locs <- schema == "*"

special_area <- which(special_locs, arr.ind = TRUE) |>
  apply(1, \(x) {
    tl <- grepl("\\d", schema[x[1] - 1, x[2] - 1])
    tr <- grepl("\\d", schema[x[1] - 1, x[2] + 1]) && !(tl && grepl("\\d", schema[x[1] - 1, x[2]]))
    l <- grepl("\\d", schema[x[1], x[2] - 1])
    r <- grepl("\\d", schema[x[1], x[2] + 1])
    bl <- grepl("\\d", schema[x[1] + 1, x[2] - 1])
    br <- grepl("\\d", schema[x[1] + 1, x[2] + 1]) && !(bl && grepl("\\d", schema[x[1] + 1, x[2]]))

    if (sum(tl, tr, l, r, bl, br) != 2) return(0)

    nums <- c()
    if (tl) {
      nums <- c(nums, sub(".*[^0-9]", "", sub(paste0("(^.{", x[2] - 2, "}\\d+).*"), "\\1", schema_lines[x[1] - 1])))
    }
    if (tr) {
      nums <- c(nums, sub(".*[^0-9]", "", sub(paste0("(^.{", x[2], "}\\d+).*"), "\\1", schema_lines[x[1] - 1])))
    }
    if (l) nums <- c(nums, sub(".*[^0-9]", "", substr(schema_lines[x[1]], 1, x[2] - 1)))
    if (r) nums <- c(nums, sub("[^0-9].*", "", substr(schema_lines[x[1]], x[2] + 1, dims[2])))
    if (bl) {
      nums <- c(nums, sub(".*[^0-9]", "", sub(paste0("(^.{", x[2] - 2, "}\\d+).*"), "\\1", schema_lines[x[1] + 1])))
    }
    if (br) {
      nums <- c(nums, sub(".*[^0-9]", "", sub(paste0("(^.{", x[2], "}\\d+).*"), "\\1", schema_lines[x[1] + 1])))
    }

    prod(as.numeric(nums))
  })

sum(special_area)
```
